# Statistics

In this chapter, we will cover the basics of statistics, including descriptive statistics, hypothesis testing, and correlation analysis. We will also discuss how to perform these analyses using R.
What is a statistic?
A statistic is a numerical value that summarizes a dataset. Statistics can be used to describe the characteristics of a dataset, test hypotheses, and make predictions. There are two main types of statistics: descriptive statistics and inferential statistics.


## Statistics on a single variable

When we talk about a "variable" in R or other analyses, we may be talking about the thing we are measuring OR the measurements themselves.  
For example, when we talk about the "age" variable, we may be talking about "age" as a general measurable thing OR about the measurements of "age" on subjects.
The same holds true when working with data and statistics. 
In the context of this chapter, when we are talking about "statistics on a single variable," we are referring to summarizing the measurements of that variable. 

Let's make this concrete by simulating 50 adult patient ages. 

```{r}
# set the random number generator seed to make results reproducible
set.seed(123)

# simulate some adult ages
age <- sample(18:99, size = 50, replace = TRUE)
```

Take a moment to examine that data. Consider using a histogram (`hist()`), printing the variable, or using `summary()`.

### Measures of central tendency

Now that we have our simple `age` variable, calculate the mean and the median.

```{r}
mean(age)
median(age)
```

These measures are called measures of central tendency and give us a sense of the "center" of the distribution. 
Note that the mean and median do not give the same result. 
Why is that?

As an exercise, try using the `quantile()` function to find the median rather than the `median()` function.

### Measures of dispersion or spread

In addition to measuring the "middle" of a variable, we often also desire a statistic to represent the "spread" of a variable.
Measures such as the standard deviation and interquartile range (IQR) help us with that. 

```{r}
sd(age)
IQR(age)
```

## Statistics on two variables

To work on this section, let's load up our insurance data.

```{r eval=FALSE}
insurance_url <-
    insurance <- readr::read_csv(insurance_url)
```



### Correlation

### Hypothesis testing

